---
title: "Reproducible Research assignment 1"
author: "Samiran Roy"
---

This is my submission for the reproducible research Peer graded assignment 1

### Getting necessary libraries
```{r, warning=FALSE,message=FALSE}
library(RCurl)
library(plyr)
library(ggplot2)
library(zoo)
```

### Loading and preprocessing the data

#### Getting and loading data

##### Source data URL
```{r}
fileURL <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
```

##### Check if source data already exists; Else Download
```{r}
if(!file.exists("./data")){ dir.create("./data") }
if(!file.exists("./data/repdata_data_activity.zip"))
  { 
    download.file(fileURL, "./data/repdata_data_activity.zip", mode="wb")
  }
unzip("./data/repdata_data_activity.zip")
```

##### Load data into dataset
```{r}
activityData <- read.csv("./data/repdata_data_activity/activity.csv")
str(activityData)
```

#### Preprocessing data to meet analysis requirements

##### Account for missing values
```{r}
sum(is.na(activityData$steps))
```

##### Create clean dataset without NA values
```{r}
activityNew <- activityData[!is.na(activityData$steps),]
str(activityNew)
```

### What is mean total number of steps taken per day?

#### Calculating the total number of steps taken each day

##### Source data contains the steps taken at 5 min intervals. To find steps per day we need to aggregate data per day

```{r}
stepsPerDay <- aggregate(activityNew$steps ~ activityNew$date, FUN=sum, )
colnames(stepsPerDay) <- c("Date","Steps")
str(stepsPerDay)
```

##### Create histogram for steps taken per day

```{r}
hist(stepsPerDay$Steps, breaks = 8, xlab = "# of Steps", main = "Steps per day", col="lightgreen")
```

#### Mean and median number of steps taken each day

```{r}
mean(stepsPerDay$Steps)
median(stepsPerDay$Steps)
```

### What is the average daily activity pattern?

#### Average the data from multiple days into 5 min intervals
```{r}
avgDailyAct <- tapply(activityNew$steps, activityNew$interval, mean)
str(avgDailyAct)
```

#### Plotting the avg number of steps against 5 min intervals throughout the day

```{r}
plot(y = avgDailyAct, x = names(avgDailyAct), type = "l", xlab = "5 Minute Interval", 
    main = "Daily Activity Pattern", ylab = "Average number of steps", col="blue")
```

#### Finding Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps

```{r}
maxInterval <- avgDailyAct[avgDailyAct==max(avgDailyAct)]
```

#### Interval 835 had the maximum number of average steps taken - 206

### Imputing missing values

#### Finding total number of missing values 

```{r}
sum(is.na(activityData$steps))
```

#### Filling missing Values - Interpolating missing NAs by using na.approx

```{r}
activityApprox <- activityData
activityApprox[which(is.na(activityApprox$steps)),1] <- 
                avgDailyAct[as.character(activityApprox[which(is.na(activityApprox$steps)),3])]
sum(is.na(activityApprox))
```

#### The missing values have been filled with the approx values from the daily average values

#### Comparing pre and post approximated values via histogram for steps per day

```{r}
approxStepsPerDay <- aggregate(activityApprox$steps ~ activityApprox$date, FUN=sum, )
colnames(approxStepsPerDay) <- c("Date","Steps")
par(mfrow=c(1,2))
hist(stepsPerDay$Steps, breaks = 8, xlab = "# of Steps", main = "Steps per day", col="lightgreen"
    , ylim=c(0,25))
abline(v = median(stepsPerDay$Steps), col = 4, lwd = 4)
hist(approxStepsPerDay$Steps, breaks = 8, xlab = "# of Steps", main = "Steps per day (missing values removed)", col="lightgreen"
     , ylim=c(0,25))
abline(v = median(approxStepsPerDay$Steps), col = 4, lwd = 4)
```

#### Calculating mean and median of dataset with missing values replaced

```{r}
mean(approxStepsPerDay$Steps)
median(approxStepsPerDay$Steps)
```

#### Finding impact of replacing missing values
```{r}
mean(approxStepsPerDay$Steps) - mean(stepsPerDay$Steps)
median(approxStepsPerDay$Steps) - median(stepsPerDay$Steps)
```

#### There is no difference in the mean however the median shifted by 1.188679

### Comparing activity patterns between weekday and weekends

#### Using weekdays function to factor the dataset into weekdays and weekends

```{r}
activityApprox$date <- as.Date(activityApprox$date) 
activityApprox$wkday <- weekdays(activityApprox$date)
activityApprox$fwkday <- as.factor(c("weekend","weekday"))
activityApprox[activityApprox$wkday == "Sunday" | activityApprox$wkday == "Saturday",5] <- factor("weekend")
activityApprox[!(activityApprox$wkday == "Sunday" | activityApprox$wkday == "Saturday"),5] <- factor("weekday")
```

#### Creating separate datasets for weekdays and weekends

```{r}
activityWeekdays <- subset(activityApprox, fwkday == "weekday")
activityWeekends <- subset(activityApprox, fwkday == "weekend")
avgDailyAct_weekday <- tapply(activityWeekdays$steps, activityWeekdays$interval, mean)
avgDailyAct_weekend <- tapply(activityWeekends$steps, activityWeekends$interval, mean)
par(mfrow=c(1,2))
plot(y = avgDailyAct_weekday, x = names(avgDailyAct_weekday), type = "l"
    , xlab = "5 Minute Interval", main = "Daily Activity Pattern (Weekday)"
    , ylab = "Average number of steps", col="blue")
plot(y = avgDailyAct_weekend, x = names(avgDailyAct_weekend), type = "l"
    , xlab = "5 Minute Interval", main = "Daily Activity Pattern (Weekend)"
    , ylab = "Average number of steps", col="blue")
```

#### The activity trends are significantly different during weekdays and weekends. 
#### During weekdays the average number of steps taken are very high towards the beginning of the day with considerably lower numbers for the rest of the day.
#### During weekends the average number of steps taken peak during the mornings however they stay high during the entire day with multiple mini peaks.